CODECHECK: An open-science initiative for the independent execution of computations underlying research articles.

### 2021-01-11 codecheck-paper.tex
resolve TODOs

\section*{TODO}\label{todo}

\begin{enumerate}
\item Check with Giuliano re: ``code works''.
\item Can we introduce "research compendium" as a term?
\item Should we make code and data sharing a prerequisite or leave that door open?
\item Take a look at the Lancet's statistical review process (areas for improvement by authors) \url{https://www.thelancet.com/journals/lancet/article/PII0140-6736(92)90409-V/fulltext}
\item How to relate to "reproducibility does not matter but trust does", see \url{https://gmbecker.github.io/MayInstituteKeynote2019/outline.html}
\item Should we ask the people currently listed in the acknowledgements?
\item revisit \url{https://github.com/codecheckers/discussion}
\item \url{https://hdsr.mitpress.mit.edu/pub/f0obb31j/release/1} and other articles form the HDSR special issue
\item \url{https://av.tib.eu/media/42484} > "Code will make systematic mistakes, not random mistakes", "code is more reliable (and auditable) than a human"
\item \url{https://doi.org/10.1016/j.isci.2020.101698} - "Rigor and Transparency Index may serve as a proxy for methodological quality" - CODECHECK doesn't evaluate that, could be used for distinction
\item Make code Feedback and leave better than found a principle, not an option? Just checking the code does not give enough credit (?), but improving it certainly does, and good peer review improves (watch out to not do full software review though)
\end{enumerate}

@@

We consider a \emph{single codechecker} to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, actually match should be a very factual 
process and not
(see TODO: check ref \ref{when-to-do-a-codecheck}), a mediator, e.g., the editor, or a
second opinion might be warranted in rare cases.

@@

Publisher staff represents the most controlled but also resource-intensive
option. Hiring the required technical and or domain expertise puts a financial
burden on the publisher, of course, but such a commitment also shows the 
value given to publishing reproducibility and would positively impact the 
submission experience, because of more controllability, and external 
perception of the journal or conference. A staff codechecker reduces the 
challenges when it comes to open communication and awarding credit, as
in-house personnel may be more readily be given access to non-anonymised
submission material and receives a proper salary instead of public credit
as part of the scientific community.

@@

\textbf{4. Workflows must be auditable.}
This requires that the codechecker 
has at least enough material to validate the workflow outputs submitted by 
the authors. Stark~\cite{stark_before_2018} calls this `preproducibility'
and the ICERM report \cite{stodden_setting_2013} defines the level
``Auditable Research''  similarly.
Note that openness is not required here, as to accommodate 
communities of practice and review workflows that (currently) do not want
to or are not able to work openly.
A completed check means that code could be executed at least once without
critical errors or
warnings using the provided instructions, and, therefore, all code and data 
was given and could be investigated more deeply or extended in the future.
Ideally, this is a "one~click" step, but achieving this requires particular 
skills. It is also hard to create a sufficient level of documentation for 
third parties. Furthermore automation may lead to people gaming the system
or reliance on technology hiding important details. All such aspects can
reduce the understandability of the material, which we estimate to be the
best mean to ensure long-term transparency and usefulness.
That is why the possibility to communicate is more important
than perfected workflows, and why a CODECHECK is not automated.
We acknowledge that others have argued in favour of bitwise reproducibility
because in the long run it can be automated\footnote{Konrad Hinsen on 
Twitter: \url{https://twitter.com/khinsen/status/1242842759733665799}};
but until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.} % or "design"?
By default, unless there are strong
reasons to the contrary (e.g., clinically-sensitive data), all code and
data, both from author and codechecker, will be made freely available, i.e., 
under open licenses, before or when the certificate is published.
The code and data  publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed as a recognition of
shortcomings in asserting reproducibility of submissions and understand the
complex situation that lead to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives and a perspective to improve education
around reproducibility of methods and communication of comprehensive 
computational workflows, with the long term goal for a specific community or
journal's target audience to reach a state where the level of verification 
and trust provided by codecheckers becomes part of peer review.

@@

\subsection*{Register}\label{register}
% count certificates: 
% curl https://codecheck.org.uk/register/stats.json
% -1 for Hathway-Goodman (no empty last line, so no -1 for the column headers needed)

We have a curated list of 22 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons and fall into three themes:
(1) `classic' papers from computational neuroscience
(2) COVID-19 modelling preprints, and
(3) GIScience.  
The first theme was the initial set of papers used to explore the concept
of CODECHECK. The idea was to take very well known historical articles from a 
specific domain, in this case the work area of co-author SJE, possibly 
as as a conversation starter for potential collaborating journals.
The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but the certificate
has then been acknowledged/cited in the paper \cite{Davies2020-vj}.
The third theme represents co-author DN's service as Reproducibility
Reviewer at the AGILE conference series.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series \cite{nust_improving_2020}.
While using slightly different terms and infrastructure, e.g., ``reproducibility
reports'' are published on OSF instead of certificates on Zenodo, the 
reproducibility reviews adhere to the codecheck principles.
Furthermore, a small number of checks was completed as part of a peer reviews
for GIScience journals.

@@

\textbf{Can't someone cheat?} Yes. We don't check it is correct code,
just simply that it runs. This `mechanical' test is indeed a very low bar.
Yet, by having the code and (raw) data openly deposited, interested parties can
then examine the code, and knowing that the code will be open we hope will
be an incentive for authors to share. It will also give the opportunity 
to future researchers, potentially with new methods, to look for errors.
This is a much more effective means to avoid cheating than an arms race
trying to catch the cheater with closed datasets and code. This approach
is comparable to the storing of blood samples in sports to detect
doping in the future (cf. \cite{everythinghertz97}).
Another picture that helped us to define the scope of a CODECHECK is:
the codechecker is the forensic photographer, making sure all the details
are captured so that an investigator may, now or in the future, scrutinize
them.

\textbf{Who's got time for more peer review?} Agree.
Ignoring the complex community contract behind volunteered peer review,
the time spent with CODECHECKs likely reduces the individual's time to work
as a reviewer.
However, there are differences to peer review.
First, the technical nature of a CODECHECK can provide a more 
clearly defined scope and expectations compared to conventional 
peer review. The authors are told what to provide and the codechecker
can be told when to stop, for example.
Second, the specific skill set allows to introduce different groups
into the peer review process. ECRs might be attracted to working with
groups to learn more about recent methods and reproducibility practices.
Research software experts (Research Software Engineers, RSE) who might
not regularly be involved in writing or reviewing papers could be 
interested to increase their connection with scholarly practices.
An extra codechecker may simplify the matchmaking an editor has to do
when identifying suitable reviewers for a submission, as technical and
topical expertise can be provided by different people, who must not even
be traditional peer reviewers. In fact, the codechecker could even be
a software professional employed by a publisher.
Third, recall that we expect CODECHECKs to be not anonymous and always
publicly deposited, which is common practice for some journals, but not
broadly practiced. The possibility to ask for clarification or more 
material/information can greatly reduce the digging required and thereby
the time spend.
We also found that the focus on the mechanics of a
workflow and the ability to interact with the authors can make
reproductions very educational. It's a different role and as such could
be a welcome alternation/diversion for researchers to give back their 
time to the community.

\textbf{Who finds the codechecker?} Good point! The journal, using either
their own team of codecheckers or the public database. Researchers
however think this is an important issue, and are volunteering.

@@

\textbf{Not revolutionary enough?}
CODECHECK's approach is to acknowledge the shortcomings around 
computational reproducibility and iteratively improve within the current
system of scholarly communication through journals.
It remains to be proved if this approach is welcomed broadly and if 
involving the traditional system of publishers helps to push the 
conversation along or not.
We have discussed stronger more drastic rules at length, such as only 
considering diamond Open Access journals, community-led no-fee journals,
or journals with collaborative/open peer review, but eventually decided
against it.
Note that the focused nature of the CODECHECK principles allows journals
and publishers to innovate on the finance model and peer review practices
at their own pace.

@@

SJE is on the editorial board at the journal \emph{Scientific Data}.
DN is reproducibility editor at the Association of Geographic
Information Laboratories in Europe's (AGILE) annual conference.

>>>>

@@

We consider a \emph{single codechecker} to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, actually match should be a very factual 
process and not
require much debate or opinion. Code usually makes systematic and repeatable
mistakes is is thereby more reliable and auditable than processes controlled
by humans \cite{tibav:42484}, e.g., in a laboratory.

Depending on the importance of the CODECHECK
(see \ref{when-to-do-a-codecheck}), a mediator, e.g., the editor, or a
second opinion might be warranted in rare cases.

@@

A publisher's \textbf{editorial staff} represents the most controlled but also
resource-intensive
option. Hiring the required technical and or domain expertise puts a financial
burden on the publisher, of course, but such a commitment also shows the 
value given to publishing reproducibility and would positively impact the 
submission experience, because of more controllability, and external 
perception of the journal or conference. A staff codechecker reduces the 
challenges when it comes to open communication and awarding credit, as
in-house personnel may be more readily be given access to non-anonymised
submission material and receives a proper salary instead of public credit
as part of the scientific community.
Keeping diverse specialists as staff to realise a codechecking process is 
unlikely to be feasible, especially for small or independent publishers.

@@

\textbf{4. Workflows must be auditable.}
This requires that the codechecker 
has at least enough material to validate the workflow outputs submitted by 
the authors. Stark~\cite{stark_before_2018} calls this `preproducibility'
and the ICERM report \cite{stodden_setting_2013} defines the level
``Auditable Research''  similarly.
Every community may establish their own good practices or adapt generic
concepts and practical tools, such as publishing all building blocks
of science in a research compendium
\footnote{\url{https://research-compendium.science/}}.
A completed check means that code could be executed at least once without
critical errors or
warnings using the provided instructions, and, therefore, all code and data 
was given and could be investigated more deeply or extended in the future.
Ideally, this is a "one~click" step, but achieving this requires particular 
skills. It is also hard to create a sufficient level of documentation for 
third parties. Furthermore automation may lead to people gaming the system
or reliance on technology hiding important details. All such aspects can
reduce the understandability of the material, which we estimate to be the
best mean to ensure long-term transparency and usefulness.
That is why the possibility to communicate is more important
than perfected workflows, and why a CODECHECK is not automated.
We acknowledge that others have argued in favour of bitwise reproducibility
because in the long run it can be automated\footnote{Konrad Hinsen on 
Twitter: \url{https://twitter.com/khinsen/status/1242842759733665799}};
but until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.} % or "design"?
By default, unless there are strong
reasons to the contrary (e.g., clinically-sensitive data), all code and
data, both from author and codechecker, will be made freely available, i.e., 
under open licenses, before or when the certificate is published.
Note that openness is not required for the paper itself as to accommodate 
journals to transition to sustainable open access models.
The code and data publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed (a) as a recognition of
shortcomings in asserting reproducibility of submissions and (b)
with an understanding of the
complex situation that lead to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives and a perspective to improve education
around reproducibility of methods and communication of comprehensive 
computational workflows, with the long term goal for a specific community or
journal's target audience to reach a state where the level of verification 
and trust provided by codecheckers becomes part of peer review and regular
merit of all accepted papers.

@@

\subsection*{Register}\label{register}

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons and fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.  
The first theme was the initial set of papers used to explore the concept
of CODECHECK. The idea was to take very well known historical articles from a 
specific domain, in this case the work area of co-author SJE, possibly 
as as a conversation starter for potential collaborating journals.
The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but the certificate
has then been acknowledged/cited in the paper \cite{Davies2020-vj}.
The third theme represents co-author DN's service as Reproducibility
Reviewer at the AGILE conference series.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series \cite{nust_improving_2020}.
While using slightly different terms and infrastructure, e.g., ``reproducibility
reports'' are published on OSF instead of certificates on Zenodo, the 
reproducibility reviews adhere to the codecheck principles.
Furthermore, a small number of checks was completed as part of a peer reviews
for GIScience journals.

@@

\textbf{Can't someone cheat?} Yes. We don't check it is correct code or
sound science,
just simply that it runs. This `mechanical' test is indeed a very low bar.
Yet, by having the code and (raw) data openly deposited, interested parties can
then examine the code, and knowing that the code will be open we hope will
be an incentive for authors to share. It will also give the opportunity 
to future researchers, potentially with new methods, to look for errors.
This is a much more effective means to avoid cheating than an arms race
trying to catch the cheater with closed datasets and code. This approach
is comparable to the storing of blood samples in sports to detect
doping in the future (cf. \cite{everythinghertz97}).
Another picture that helped us to define the scope of a CODECHECK is:
the codechecker is the forensic photographer, making sure all the details
are captured so that an investigator may, now or in the future, scrutinize
them.

\textbf{Who's got time for more peer review?} Agree.
Ignoring the complex community contract behind volunteered peer review,
the time spent with CODECHECKs likely reduces the individual's time to work
as a reviewer.
However, there are differences to peer review.
First, the technical nature of a CODECHECK can provide a more 
clearly defined scope and expectations compared to conventional 
peer review. The authors are told what to provide and the codechecker
can be told when to stop, for example.
Second, the specific skill set allows to introduce different groups
into the peer review process. ECRs might be attracted to working with
groups to learn more about recent methods and reproducibility practices.
Research software experts (Research Software Engineers, RSE) who might
not regularly be involved in writing or reviewing papers could be 
interested to increase their connection with scholarly practices.
An extra codechecker may simplify the matchmaking an editor has to do
when identifying suitable reviewers for a submission, as technical and
topical expertise can be provided by different people, who must not even
be traditional peer reviewers. In fact, the codechecker could even be
a software professional employed by a publisher.
Third, recall that we expect CODECHECKs to be not anonymous and always
publicly deposited, which is common practice for some journals, but not
broadly practiced. The possibility to ask for clarification or more 
material/information can greatly reduce the digging required and thereby
the time spend.
We also found that the focus on the mechanics of a
workflow and the ability to interact with the authors can make
reproductions very educational. It's a different role and as such could
be a welcome alternation/diversion for researchers to give back their 
time to the community.
With code and workflows, there also is a good chance that a codechecker's
feedback can have direct impact and lead to improvements of the author's 
work that comes with a rewarding experience or even tangible contributions.
While such impact is also part of idealistic peer review, it is
rather rare beyond paraphrased anonymous acknowledgement.

@@

\textbf{Not revolutionary enough?}
CODECHECK's approach is to acknowledge the shortcomings around 
computational reproducibility and iteratively improve within the current
system of scholarly communication through journals.
It remains to be proved if this approach is welcomed broadly and if 
involving the traditional system of publishers helps to push the 
conversation along or not.
We have discussed stronger more drastic rules at length, such as only 
considering diamond Open Access journals, community-led no-fee journals,
or journals with collaborative/open peer review, but eventually decided
against them.
We also have considered to require latest technologial solutions which can
support reproducibility (cf.~\cite{konkol_publishing_2020}), but have 
decided to focus on the human interface and the judgement of an experienced
researchers/developer as a more sustainable approach.
The structurelessness of the required data and software gives the needed
freedom to enable all types of research. It could be complemented by
automated scoring (e.g.,~\cite{menke_rigor_2020}, yet automation and 
metrics bears risks that lead us to put our trust in the codechecker.
Note that the focused nature of the CODECHECK principles allows journals
and publishers to innovate on the financial model and peer review practices
at their own pace.

@@

SJE is on the editorial board at the journal \emph{Scientific Data}.
DN is reproducibility chair at the Association of Geographic
Information Laboratories in Europe's (AGILE) annual conference.
